---
format: 
   revealjs:
      theme: ["../theme/q-theme.scss"]
      slide-number: c/t
      logo: POPUHEALSMPH_color-flush.png
      code-copy: true
      center-title-slide: false
      code-link: true
      code-overflow: wrap
      highlight-style: a11y
      height: 1080
      width: 1920
      chalkboard: true
execute: 
   eval: true
   echo: true
---
```{r load libraries, echo=F}
library(tidyverse)
library(ggdag)
library(dagitty)
library(gridExtra)
```
<h1> Lecture 1: Correlated data </h1>

<h2> PHS 651: Advanced regression methods </h2>

<hr>

<h3> Mary Ryan, PhD </h3>

<h3> September 10, 2024 </h3>

---

## What is correlated data?

- [Independent data]{.underline}: data where each observation of a variable are not meaningfully correlated with other observations of the same variable

- [Correlated data]{.underline}: data where each observation of a variable is not entirely independent of other observations of the same variable
   - AKA: multilevel/hierarchical data

---

## Why is data correlated?

- Natural group membership

   - Observations belonging to one group more similar than observations belonging to another group
   
   - e.g., Household members, classroom students, eyes in a person's head
   
- Clustered by design
   
   - Exposure applied to entire group based on single characteristic
   
      - e.g., All residents of City X are exposed to a new policy
   
      - Exposure can be randomized or not
   
   - Cluster/multi-stage sampling
      
      - e.g., NHANES

---

## How does clustering affect outcomes?

```{r cluster DAG, echo=F, fig.align="center"}
straight <- dagify(
  E ~ C,
  O ~ E,
  coords = list(x = c(C = 1, E = 2, O = 3), y = c(C = 1, E = 1, O = 1))
) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag() +
  ylim(0, 2) +
  geom_dag_edges_arc(curvature = c(0,0))

med <- dagify(
  O ~ C,
  E ~ C,
  O ~ E,
  coords = list(x = c(C = 1, E = 2, O = 3), y = c(C = 1, E = 1, O = 1))
) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag() +
  ylim(0, 2) +
  geom_dag_edges_arc(curvature = c(0,0.5, 0))

precision <- dagify(
  O ~ E,
  O ~ C,
  coords = list(x = c(E = 1, O = 2, C=2.5), y = c(E = 1, O = 1, C=0.5))
) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point() +
  geom_dag_text() +
  theme_dag() +
  ylim(-0.5, 2) +
  geom_dag_edges_diagonal(curvature = c(0,0))

grid.arrange(straight, med, ncol=2)
precision
```

Clustering is both a [data structure]{.underline} element and potentially a [confounding variable]{.underline}

<!-- Seaman S, Pavlou M, Copas A. Review of methods for handling confounding by cluster and informative cluster size in clustered data. Stat Med. 2014 Dec 30;33(30):5371-87. doi: 10.1002/sim.6277. Epub 2014 Aug 4. PMID: 25087978; PMCID: PMC4320764. -->

---

## Other issues from correlated data

- Groups/clusters may not be the same size

- "Clustering" of data may also affect variance

   - Some groups have lower variation than others

   - This breaks the [homoskedasticity]{.underline} assumption of OLS

      - Recall PHS 552!
      
---

## Measuring "clustered-ness"

- [Intracluster correlation coefficient]{.underline} (ICC): measure of within- vs. between-cluster variation

   - Cluster-specific mean: $\bar{y}_i = \frac{1}{n_i} \sum_{j=1}^{n_i} y_{ij}$
   
   - Variation in cluster means: $\gamma^2 = \text{var}(\bar{y}_i)$
   
   - Individual-level variation: $\sigma^2 = \text{var}(y_{ij})$
   
   - ICC: $\rho = \frac{\gamma^2}{\gamma^2 + \sigma^2}$