---
format: 
   revealjs:
      theme: ["../theme/q-theme.scss"]
      slide-number: c/t
      logo: POPUHEALSMPH_color-flush.png
      code-copy: true
      center-title-slide: false
      code-link: true
      code-overflow: wrap
      highlight-style: a11y
      height: 1080
      width: 1920
      chalkboard: true
      from: markdown+emoji
      fragment: true
      auto-stretch: false
execute: 
   eval: true
   echo: true
editor: 
  markdown: 
    wrap: 72
---

```{r load libraries, echo=F}
library(tidyverse)
library(ggdag)
library(dagitty)
library(gridExtra)
library(qrcode)
```

<h1>Lecture 7: Modeling the (longitudinal) covariance</h1>

<h2>PHS 651: Advanced regression methods</h2>

<hr>

<h3>Mary Ryan Baumann, PhD</h3>

<h3>October 22, 2024</h3>

::: {.absolute bottom=-100 left=260 width=1500}

Slides found at: <https://maryryan.github.io/PHS-651-slides/PHS-651-7/slides-7>

:::

::: {.absolute bottom=-150 left=100 width=150}
```{r qr code, echo=F, fig.height=2, fig.width=2}
plot(qr_code("https://maryryan.github.io/PHS-651-slides/PHS-651-7/slides-7"))
```
:::

---

## Zoom recording disclaimer

<br>

This class is being conducted in person, as well as over [Zoom]{.alert}. As the instructor, I will be [recording]{.alert} this session. I have disabled the recording feature for others so that no one else will be able to record this session. I will be posting this session to the course’s website.

<br>

If you have privacy concerns and do not wish to appear in the recording, you may turn video off (click “stop video”) so that Zoom does not record you.

<br>

The chat box is always open for discussion and questions to the entire class. You may also send messages privately to the instructor. Please note that Zoom saves all chat transcripts.

---

## Longitudinal covariance/correlation structures

Previously we lightly touched on correlation structures for clustered, non-longitudinal data

- Remember: independence, exchangeable, unstructured from Lectures 2 and 4

. . .

With longitudinal data, there are a much wider range of potential structures

- We'll explore some of those today

---

## Unstructured covariance

---

## Covariance Pattern Models

When attempting to impose some structure on the covariance, a subtle balance needs to be struck

- Too little structure means too many parameters means imprecise estimate (large variance)

- Too much structure means more opportunity for mis-specification means inaccurate estimate (large bias)

   - Classic tradeoff between bias and variance
   
Covariance pattern models have their basis in models for serial correlation originally developed for time series data

---

## Exchangeable correlation/compound symmetry

[Compound symmetry/exchangeable correlation]{.alert} assumes:

- Correlation is the same across all observations in the cluster

- [We saw this one before in lecture 2 and 4!]{.alert}

$$\begin{bmatrix}
1 & \alpha & \dots & \alpha\\
\alpha & \ddots & \dots & \alpha\\
\vdots & \dots & \ddots & \vdots\\
\alpha & \dots & \alpha & 1
\end{bmatrix}$$

You'll always be estimating a single correlation $\alpha$, regardless of $n$

. . .

This pattern has strong assumptions (measurements are exchangeable)

- [This is usually not valid with longitudinal data]{.alert}

---

## Toeplitz correlation

[Toeplitz]{.alert} assumes:

- Correlation is different between each observation

   - The correlation between two observations depends only on the [number of time points]{.alert} separating them

$$\begin{bmatrix}
1 & \alpha_1 & \dots & \alpha_{n-1}\\
\alpha_1 & \ddots & \dots & \alpha_{n-2}\\
\vdots & \dots & \ddots & \vdots\\
\alpha_{n-1} & \dots & \alpha_1 & 1
\end{bmatrix}$$

You'll be estimating $(n-1)$ correlations

. . .

[Only appropriate when measurements are (approximately) equally spaced!!]{.alert}


---

## Autoregressive correlation

An [autoregressive]{.alert} structure assumes:

- Correlation gradually decreases as the distance between observations grows

$$\begin{bmatrix}
1 & \alpha^1 & \dots & \alpha^{n-1}\\
\alpha^1 & \ddots & \dots & \alpha_{n-2}\\
\vdots & \dots & \ddots & \vdots\\
\alpha^{n-1} & \dots & \alpha^1 & 1
\end{bmatrix}$$

You'll always be estimating a single correlation $\alpha$, regardless of $n$

. . .

Means less correlation for measurements taken farther apart; but [only appropriate when the measurements are (approximately) equally spaced]{.alert}

---

## Unevenly timed observations

All the structures we've looked at so far have assumed that our observed times points are approximately equally spaced

- This is appropriate when observations are things like hours, days, weeks, months, years...

. . .

... But it's not uncommon to have observation time points that **aren't** always evenly spaced!

- Think of a study where you start out with weekly observations, then move to monthly, then annual

. . .

How do we handle this?

----

## Exponential correlation

When measurement occasions are not equally-spaced over time, we can the autoregressive model can be generalized

- Let $\{t_{i1}, t_{i2}, \dots, t_{in_i}\}$ denote the observation times for the $i$th individual

. . .

- The correlations are given by:
$$Corr(Y_{ij}, Y_{ik}) = \alpha^{|t_{ij} - t_{ik}|}$$

   - Instead of looking at the how far apart the observation numbers are (e.g., visit 1 vs visit 2), it [looks at the actual distance between the observation times]{.alert}

This structure is invariant under linear transformation of the time scale

Unlike the Toeplitz and autoregression structures, the exponential structure can easily handle unbalanced data because it is based on [time as a continuum]{.alert}

---

## Banded correlation

All of the previous structures have have required us to estimate some parameter for the off-diagonal matrix elements

- What if we know at some point the correlation between observations will go to 0? [Enter the banded structure!]{.alert .fragment}

. . .

The banded structure assumes correlation is 0 beyond some specified interval

It is possible to apply a banded pattern to any of the correlation pattern models considered so far

::: {.r-stack}

::: {.fragment .fade-in-then-out}
A banded Toeplitz structure with a band size of 3 looks like:

$$\begin{bmatrix}
1 & \alpha_1 & \alpha_2 & 0 & \dots & 0\\
\alpha_1 & 1 & \alpha_1 & \ddots & \dots & 0\\
\alpha_2 & \alpha_1 & 1 &\ddots & \dots & 0\\
0 & \alpha_2 & \alpha_1& \ddots & \dots & 0\\
\vdots && \dots && \ddots & \vdots\\
0 && \dots &\alpha_2 & \alpha_1 & 1
\end{bmatrix}$$

:::

::: {.fragment .fade-in-then-out}

A banded autoregression structure with a band size of 4 looks like:
$$\begin{bmatrix}
1 & \alpha^1 & \alpha^2 & \alpha^3 & 0 & \dots & 0\\
\alpha^1 & 1 & \alpha^1 & \alpha^2& \ddots & \dots & 0\\
\alpha^2 & \alpha^1 & 1 & \alpha^1&\ddots & \dots & 0\\
\alpha^3 & \alpha^2 & \alpha^1 & 1& \ddots & \dots & 0\\
\vdots && \dots && \ddots && \vdots\\
0 && \dots & \alpha^3&\alpha^2 & \alpha^1 & 1
\end{bmatrix}$$

:::

::: {.fragment .fade-in}

A banded exponential structure with a band size of 4 looks like:
$$\begin{bmatrix}
1 & \alpha^{|t_{i1}-t_{i2}|} & \alpha^{|t_{i1}-t_{i3}|} & \alpha^{|t_{i1}-t_{i4}|} & 0 & \dots & 0\\
\alpha^{|t_{i1}-t_{i2}|} & 1 & \alpha^{|t_{i1}-t_{i2}|} & \alpha^2& \ddots & \dots & 0\\
\alpha^{|t_{i1}-t_{i3}|} & \alpha^{|t_{i1}-t_{i2}|} & 1 & \alpha^{|t_{i1}-t_{i2}|} &\ddots & \dots & 0\\
\alpha^{|t_{i1}-t_{i4}|} & \alpha^{|t_{i1}-t_{i3}|} & \alpha^{|t_{i1}-t_{i2}|} & 1& \ddots & \dots & 0\\
\vdots && \dots && \ddots && \vdots\\
0 && \dots & \alpha^{|t_{i1}-t_{i4}|}&\alpha^{|t_{i1}-t_{i3}|} & \alpha^{|t_{i1}-t_{i2}|} & 1
\end{bmatrix}$$

:::

:::

---

## Choices, choices

With way more structures to choose from, it may be more difficult to make a decision than it was with non-longitudinal data


---

## Variograms

Variogram plots are one way to assess the how strong the 
```{r cd4, echo=F}
cd4 <- read.csv("~/Desktop/teaching/PHS-651/data/Gelman-data/cd4/allvar.csv",header=T)

cd4 <- cd4 %>% 
   as.data.frame() %>% 
   dplyr::filter(treatmnt==1) %>% 
   rename(age_baseline = baseage,
          age = visage) %>% 
   mutate(time = age - age_baseline) %>% 
   select(!(treatmnt))

```

```{r variogram, fig.align='center', fig.height=8.5, fig.width=15}
library(joineR)
library(splines)

# fit outcome against time with some splines #
fit <- lm( sqrt(CD4PCT) ~ ns( time, knots=c(0.5, 1, 1.5)),
           data=cd4)

# find the residuals #
resids <- residuals( fit )

# calculate the variogram #
vario <- variogram( indv = cd4$newpid,
                    time = cd4$time,
                    Y = resids )

# only take the complete cases #
vario$svar <- vario$svar[complete.cases(vario$svar),]

# plot variogram #
plot( vario$svar[,1], vario$svar[,2], pch=".", ylim=c(0, 1.2*var(resids)),
      cex=4, cex.axis=2, cex.lab=2)

# smooth line of variogram over time #
lines( smooth.spline(vario$svar[,1], vario$svar[,2],df=3), lwd=3 )

# show where the total variation is #
abline(h=var(resids), lty=2, lwd=3)

```

