---
format: 
   revealjs:
      theme: ["../theme/q-theme.scss"]
      slide-number: c/t
      logo: POPUHEALSMPH_color-flush.png
      code-copy: true
      center-title-slide: false
      code-link: true
      code-overflow: wrap
      highlight-style: a11y
      height: 1080
      width: 1920
      chalkboard: true
      from: markdown+emoji
      fragment: true
      auto-stretch: false
execute: 
   eval: true
   echo: true
editor: 
  markdown: 
    wrap: 72
---

```{r load libraries, echo=F}
library(tidyverse)
library(ggdag)
library(dagitty)
library(gridExtra)
library(qrcode)
```

<h1>Lecture 8: Longitudinal GLMMs & GEEs</h1>

<h2>PHS 651: Advanced regression methods</h2>

<hr>

<h3>Mary Ryan Baumann, PhD</h3>

<h3>October 29, 2024</h3>

::: {.absolute bottom=0 left=260 width=1500}

Slides found at: <https://maryryan.github.io/PHS-651-slides/PHS-651-8/slides-8>

:::

::: {.absolute bottom=-50 left=100 width=150}
```{r qr code, echo=F, fig.height=2, fig.width=2}
plot(qr_code("https://maryryan.github.io/PHS-651-slides/PHS-651-8/slides-8"))
```
:::

---

## Zoom recording disclaimer

<br>

This class is being conducted in person, as well as over [Zoom]{.alert}. As the instructor, I will be [recording]{.alert} this session. I have disabled the recording feature for others so that no one else will be able to record this session. I will be posting this session to the course’s website.

<br>

If you have privacy concerns and do not wish to appear in the recording, you may turn video off (click “stop video”) so that Zoom does not record you.

<br>

The chat box is always open for discussion and questions to the entire class. You may also send messages privately to the instructor. Please note that Zoom saves all chat transcripts.

---

## Recap: Non-longitudinal GLMMs

Recall the random effects model we used for non-longitudinal clustered data:
$$g(\vec{Y}_i) = \boldsymbol{X}_i\vec{\beta} + \vec{Z}_ib_{0i}+\vec{\epsilon}_i$$

- Where we assumed that each cluster had it's own cluster-specific mean ($\beta_0 + b_{0i}$) that varied by some $b_{0i}$ around the overall sample mean $\beta_0$

- We called $b_{i0}$ the [random intercept]{.alert}

. . .

The random intercept induced an exchangeable covariance matrix:
$$\begin{align*}Var[\vec{Y}_i] &= [\text{cluster-to-cluster variation}] + [\text{within-cluster member-to-member variation}]\\
&=\begin{bmatrix}
\sigma^2_b & \sigma^2_b & \dots & \sigma^2_b\\
\sigma^2_b & \ddots & \dots & \sigma^2_b\\
\vdots & \dots & \ddots & \vdots\\
\sigma^2_b & \dots & \sigma^2_b & \sigma^2_b
\end{bmatrix} + \begin{bmatrix}
\sigma^2_\epsilon & 0 & \dots & 0\\
0 & \ddots & \dots & 0\\
\vdots & \dots & \ddots & \vdots\\
0& \dots & 0 & \sigma^2_\epsilon
\end{bmatrix}\\
&=\begin{bmatrix}
\sigma^2_b + \sigma^2_\epsilon & \sigma^2_b & \dots & \sigma^2_b\\
\sigma^2_b & \ddots & \dots & \sigma^2_b\\
\vdots & \dots & \ddots & \vdots\\
\sigma^2_b & \dots & \sigma^2_b & \sigma^2_b + \sigma^2_\epsilon
\end{bmatrix}\end{align*}$$

---

## Longitudinal GLMMs

Recall the random effects model we used for non-longitudinal clustered data:
$$g(\vec{Y}_i) = \boldsymbol{X}_i\vec{\beta} + \vec{Z}_ib_{0i}+\vec{\epsilon}_i$$

The random intercept induced an exchangeable covariance matrix:
$$Var[\vec{Y}_i] =\begin{bmatrix}
\sigma^2_b + \sigma^2_\epsilon & \sigma^2_b & \dots & \sigma^2_b\\
\sigma^2_b & \ddots & \dots & \sigma^2_b\\
\vdots & \dots & \ddots & \vdots\\
\sigma^2_b & \dots & \sigma^2_b & \sigma^2_b + \sigma^2_\epsilon
\end{bmatrix}$$

This assumes that every member of cluster $i$ has the same relationship to one another (same correlation)

- In longitudinal data, though, cluster $i$ is a single person and the "members" of the cluster are observations collected at separate time points

- An observation collected at time 1 may relate to observations at times 2 and 3 differently

. . .

One solution would be to allow for each cluster to have variation in its slope as well

. . .

- [Random slopes and intercepts!]{.alert}

---

## Random slope and intercept model

Consider a model with intercepts and slopes that vary randomly among individuals:

$$g(Y_{ij}) = \beta_0 + t_{ij}\beta_1 + W_{ij}\beta_2 + \vec{Z}_{i0}b_{i0} + t_{ij}b_{i1} + e_{ij}$$

- This model suggests that individuals vary not only in their baseline level of response (when $t_{i1} = 0$), but also in terms of their changes in the response over time

. . .

Here we assume:

$b_{i0} \sim N(0, \sigma^2_0)$
$b_{i1} \sim N(0, \sigma^2_1)$
$\epsilon_{ij} \sim N(0, \sigma^2_\epsilon)$

. . .

We also assume:
$$Cov[b_{i0}, b_{i1}] = \sigma^2_{01}$$

- The random slopes and intercepts aren't independent!

---

## Variance/covariance of the outcome

What does this mean for the covariance matrix of the outcome?

::: {.r-stack}

::: {.fragment .fade-in-then-out}

Then the variance of a single outcome measure is:
$$\begin{align*}Var[Y_{ij}] &= Var[\boldsymbol{X}_i\vec{\beta} + \boldsymbol{Z}_i\vec{b}_{i}+\epsilon_{ij}]\\
&= Var[\boldsymbol{Z}_i \vec{b}_{i}+\epsilon_{ij}]\\
&= Var[\vec{Z}_{1i} b_{i0} + t_{ij} b_{i1} +\epsilon_{ij}]\\
&= Var[\vec{Z}_{1i} b_{i0}] + Var[t_{ij} b_{i1}] + 2Cov[\vec{Z}_{1i} b_{i0}, t_{ij} b_{i1}]+Var[\epsilon_{ij}]\\
&= \sigma^2_0 + t^2_{ij}\sigma^2_1+ 2t_{ij}\sigma^2_{01} + \sigma^2_{\epsilon}\end{align*}$$

:::

::: {.fragment .fade-in}

Then the variance of a single outcome measure is:
$$\begin{align*}Var[Y_{ij}] &\color{lightgray}{= Var[\boldsymbol{X}_i\vec{\beta} + \boldsymbol{Z}_i\vec{b}_{i}+\epsilon_{ij}]}\\
&\color{lightgray}{= Var[\boldsymbol{Z}_i \vec{b}_{i}+\epsilon_{ij}]}\\
&\color{lightgray}{= Var[\vec{Z}_{1i} b_{i0} + t_{ij} b_{i1} +\epsilon_{ij}]}\\
&\color{lightgray}{= Var[\vec{Z}_{1i} b_{i0}] + Var[t_{ij} b_{i1}] + 2Cov[\vec{Z}_{1i} b_{i0}, t_{ij} b_{i1}]+Var[\epsilon_{ij}]}\\
&= \sigma^2_0 + t^2_{ij}\sigma^2_1+ 2t_{ij}\sigma^2_{01} + \sigma^2_{\epsilon}\end{align*}$$

:::

:::

::: {.r-stack}

::: {.fragment .fade-in-then-out}

And the covariance between 2 outcome measures in the same cluster is:
$$\begin{align*}Cov[Y_{ij}, Y_{ik}] &= Cov[\boldsymbol{X}_i\vec{\beta} + \boldsymbol{Z}_i\vec{b}_{i}+\epsilon_{ij}, \boldsymbol{X}_i\vec{\beta} + \boldsymbol{Z}_i\vec{b}_{i}+\epsilon_{ik}]\\
&= Cov[\vec{Z}_{1i}b_{i0} +t_{ij}b_{i1}+\epsilon_{ij}, \vec{Z}_{1i}b_{i0}+t_{ik}b_{i1}+\epsilon_{ik}]\\
&= Cov[\vec{Z}_{1i}b_{i0}, \vec{Z}_{1i}b_{i0}] + Cov[\vec{Z}_{1i}b_{i0}, t_{ik}b_{01}] + Cov[\vec{Z}_{i1}b_{i0}, \epsilon_{ik}] + Cov[t_{ij}b_{i1}, \vec{Z}_{1i}b_{i0}]\\
&~~~~~~+ Cov[t_{ij}b_{i1}, t_{ik}b_{i1}] + Cov[t_{ij}b_{i1}, \epsilon_{ik}] + Cov[\epsilon_{ij}, \vec{Z}_{1i}b_{i0}] + Cov[\epsilon_{ij}, t_{ik}b_{i1}] + Cov[\epsilon_{ij}, \epsilon_{ik}]\\
&= \sigma^2_0 + t_{ik}\sigma^2_{01} + 0 + t_{ij}\sigma^2_{01} + t_{ij}t_{ik}\sigma^2_2 + 0 + 0 + 0 + 0\\
&= \sigma^2_0 + (t_{ij} + t_{ik})\sigma^2_{01} + t_{ij}t_{ik}\sigma^2_1\end{align*}$$

:::

::: {.fragment .fade-in}

And the covariance between 2 outcome measures in the same cluster is:
$$\begin{align*}Cov[Y_{ij}, Y_{ik}] &\color{lightgray}{= Cov[\boldsymbol{X}_i\vec{\beta} + \boldsymbol{Z}_i\vec{b}_{i}+\epsilon_{ij}, \boldsymbol{X}_i\vec{\beta} + \boldsymbol{Z}_i\vec{b}_{i}+\epsilon_{ik}]}\\
&\color{lightgray}{= Cov[\vec{Z}_{1i}b_{i0} +t_{ij}b_{i1}+\epsilon_{ij}, \vec{Z}_{1i}b_{i0}+t_{ik}b_{i1}+\epsilon_{ik}]}\\
&\color{lightgray}{= Cov[\vec{Z}_{1i}b_{i0}, \vec{Z}_{1i}b_{i0}] + Cov[\vec{Z}_{1i}b_{i0}, t_{ik}b_{01}] + Cov[\vec{Z}_{i1}b_{i0}, \epsilon_{ik}] + Cov[t_{ij}b_{i1}, \vec{Z}_{1i}b_{i0}]}\\
&\color{lightgray}{~~~~~~+ Cov[t_{ij}b_{i1}, t_{ik}b_{i1}] + Cov[t_{ij}b_{i1}, \epsilon_{ik}] + Cov[\epsilon_{ij}, \vec{Z}_{1i}b_{i0}] + Cov[\epsilon_{ij}, t_{ik}b_{i1}] + Cov[\epsilon_{ij}, \epsilon_{ik}]}\\
&\color{lightgray}{= \sigma^2_0 + t_{ik}\sigma^2_{01} + 0 + t_{ij}\sigma^2_{01} + t_{ij}t_{ik}\sigma^2_2 + 0 + 0 + 0 + 0}\\
&= \sigma^2_0 + (t_{ij} + t_{ik})\sigma^2_{01} + t_{ij}t_{ik}\sigma^2_1\end{align*}$$

:::

:::

::: {.fragment .alert .absolute top=200 right=200 width="500" height="100"}
The variance and covariance are both functions of time!
:::
